---
title: "final_project"
output: html_document
date: "2023-12-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Charles Shaviro

# STATS 209

# Do Mandatory Financial Literacy Classes Cause Better Financial Outcomes?


First we import libraries and set our seed to 209.

```{r}
set.seed(209)
library(tidyverse)
library(magrittr)
library(DOS2)
library(optmatch)
library(RItools)
source("utility.R")
```

Summarize.match function from utility.R puts the results of a pair match
in a nice form. Usage: summarize.match(dataset,pairmatch_output)

```{r}
summarize.match <- function(dat, ms, ps.name="prop", keep.mset=FALSE) {
  adat <- dat
  adat$mset <- ms
  adat <- adat[!is.na(adat$mset),]
  adat.treat <- adat[adat$Z==1, ]
  adat.ctrl <- adat[adat$Z==0, ]
  
  adat.m <- merge(adat.treat, adat.ctrl, by="mset", suffixes=c(".1", ".0"))
  
  if(!keep.mset) {
    adat.m <- adat.m[, -which(names(adat.m) %in% c("Z.1", "Z.0", "mset"))]
  } else {
    adat.m <- adat.m[, -which(names(adat.m) %in% c("Z.1", "Z.0"))]        
  }
  adat.m <- adat.m[, sort(names(adat.m), index.return=TRUE)$ix]
  
  p0.name <- paste0(ps.name,".", 0)
  p1.name <- paste0(ps.name,".",1)
  
  adat.m.tmp.1 <- adat.m[, -which(names(adat.m) %in% c(p0.name, p1.name))]
  adat.m.tmp.2 <- adat.m[, c(p0.name, p1.name)]
  
  adat.m <- cbind(adat.m.tmp.1, adat.m.tmp.2)
  
  return(adat.m)
}

help(dynarski)

data(dynarski)

dynarski <- dynarski[,-1]

set.seed(209)
```






## Implementing 1:1 Matching 

```{r}
filename = "fin_lit_data_treat=all_scale=True.csv"
df = read.csv(filename)
print(colnames(df))
```

```{r}
covariate_balance <- plot(xBalance(Z ~ RACE_ETHNICITY + HIGHEST_EDUCATION_OF_RAISERS + BINARIZED_GENDER + LAYOFF_PANDEMIC + STATE + EDUCATION_LEVEL + NUM_DEPENDENT_CHILDREN + AGE + EXPECT_INHERIT_10K_PLUS, data=df))

#ggsave("other_Treat=ALL_covariate_balance.png", covariate_balance, width = 10, height = 6, units = "in", dpi = 300)


```


```{r}
ggplot(df,aes(x=EDUCATION_LEVEL,fill=factor(Z))) +
  geom_density(alpha=0.5) +
  theme_classic() +
  xlab("Education Level") +
  scale_fill_discrete(name = "Z (Fin Literacy Class in HS)")
```



```{r}
# ggplot(dynarski,aes(x=factor(edm, levels=c(1,2,3,4),
#     labels=c("<high school","high school", "some college","BA or higher")),
#             fill=factor(zb),
#             group=factor(zb),
#             y=..prop..)) +
#   geom_bar(position=position_dodge()) +
#   theme_classic() +
#   xlab("Mother's Education") +
#   ylab("proportion") +
#   scale_fill_discrete(name = "Z (Fin Literacy Class in HS)")

```

### We compute propensity scores from all of our covariates

```{r}

df$prop <- glm(Z ~ RACE_ETHNICITY + HIGHEST_EDUCATION_OF_RAISERS + BINARIZED_GENDER + LAYOFF_PANDEMIC + STATE + EDUCATION_LEVEL + NUM_DEPENDENT_CHILDREN + AGE + EXPECT_INHERIT_10K_PLUS,
family=binomial, data=df)$fitted.values

print(length(df$prop))

```

### And we plot our propensity scores


```{r}

prop_score_plot <- ggplot(df,aes(x=prop,fill=factor(Z))) +
geom_density(alpha=0.5) +
theme_classic() +
xlab("Propensity Score") +
scale_fill_discrete(name = "Z (Fin. Literacy Class in HS)")

#ggsave("Treat=ALL_0.1_prop_score_density.png", prop_score_plot, width = 10, height = 6, units = "in", dpi = 300)

```

## Matching

We conduct multivariate matching on all variables, with the Mahalanobis distance (and no caliper),
obtain a matched-pair set. Using the ‘xBalance’ function, plot the covariate balance before and after
matching.


```{r}
options("optmatch_max_problem_size" = Inf)
match.1 <- smahal(df$Z, df[,c("RACE_ETHNICITY", "HIGHEST_EDUCATION_OF_RAISERS", "BINARIZED_GENDER", "LAYOFF_PANDEMIC", "STATE", "EDUCATION_LEVEL","NUM_DEPENDENT_CHILDREN","AGE","EXPECT_INHERIT_10K_PLUS")])
ms.1 <- pairmatch(match.1, data=df)
plot(xBalance(Z ~ RACE_ETHNICITY + HIGHEST_EDUCATION_OF_RAISERS + BINARIZED_GENDER + LAYOFF_PANDEMIC + STATE + EDUCATION_LEVEL + NUM_DEPENDENT_CHILDREN + AGE + EXPECT_INHERIT_10K_PLUS + prop -1, strata=list(unstrat=NULL, ms.1=~ms.1), data=df), ggplot=TRUE)

```


We see that our matching has significantly improved covariate imbalance.

Let's see if it will improve further with caliper matching.


```{r}

match.2 <- addcaliper(match.1, z=df$Z, p=df$prop, caliper=1)
ms.2 <- pairmatch(match.2, data=df)
print(ms.2, grouped = TRUE)



balance_plot <- plot(xBalance(Z ~ RACE_ETHNICITY + HIGHEST_EDUCATION_OF_RAISERS + BINARIZED_GENDER + LAYOFF_PANDEMIC + STATE + EDUCATION_LEVEL + NUM_DEPENDENT_CHILDREN + AGE + EXPECT_INHERIT_10K_PLUS + prop -1, strata=list(unstrat=NULL, ms.2=~ms.2), data=df), ggplot=TRUE)

# Save the plot to a PNG file
#ggsave("Treat=ALL_UNscaled_imbalance_plot_0.05.png", balance_plot, width = 10, height = 6, units = "in", dpi = 300)


```


```{r}
summary_2 <- summarize.match(df, ms.2)
print(summary_2)

pairs_avg_diff <- mean(abs(summary_2$prop.0 - summary_2$prop.1))

pairs_max_diff <- max(abs(summary_2$prop.0 - summary_2$prop.1))

sprintf("The average absolute difference in propensity scores is: %f",pairs_avg_diff)
sprintf("The maximum absolute difference in propensity scores is: %f",pairs_max_diff)



sprintf("Length of summary table: %f",(length(summary_2)) )
 
test_stat <- mean((summary_2$FIN_HEALTH.1 - summary_2$FIN_HEALTH.0))
print(test_stat)
 
n <- length(summary_2$FIN_HEALTH.1)
print(n)
 
count <- 0

for (i in 1:1000) {
  zi_s <- rbinom(n, 1, p=0.5)
  local_test_stat <- mean((2*zi_s - 1)*(summary_2$FIN_HEALTH.1-summary_2$FIN_HEALTH.0))
  if (local_test_stat >= test_stat) {
    count <- count + 1
  }
}

print(count)
pval <- count/1000.0

sprintf("The P-Value is: %f",pval)



```

Now we compute a bias-corrected estimate of the average treatment effect on and
estimate its variance.

```{r}
columns_0 = c("RACE_ETHNICITY.0", "HIGHEST_EDUCATION_OF_RAISERS.0", "BINARIZED_GENDER.0","LAYOFF_PANDEMIC.0", "STATE.0", "EDUCATION_LEVEL.0", "NUM_DEPENDENT_CHILDREN.0", "AGE.0", "EXPECT_INHERIT_10K_PLUS.0", "FIN_HEALTH.0", "FIN_HEALTH.1")
columns_1 = c("RACE_ETHNICITY.1", "HIGHEST_EDUCATION_OF_RAISERS.1", "BINARIZED_GENDER.1", "LAYOFF_PANDEMIC.1", "STATE.1", "EDUCATION_LEVEL.1", "NUM_DEPENDENT_CHILDREN.1", "AGE.1", "EXPECT_INHERIT_10K_PLUS.1", "FIN_HEALTH.0", "FIN_HEALTH.1")

controls <- summary_2[, columns_0]
treats <- summary_2[, columns_1]


columns_0 = c('incmiss.0', 'afqtpct.0', 'black.0', 'edm.0', 'edmissm.0', 
                'faminc.0', 'female.0', 'hisp.0', 'y.0', 'y.1')
columns_1 = c('incmiss.1', 'afqtpct.1', 'black.1', 'edm.1', 'edmissm.1', 
              'faminc.1', 'female.1', 'hisp.1','y.0','y.1')


colnames(controls) <- c("RACE_ETHNICITY", "HIGHEST_EDUCATION_OF_RAISERS", "BINARIZED_GENDER","LAYOFF_PANDEMIC", "STATE", "EDUCATION_LEVEL", "NUM_DEPENDENT_CHILDREN", "AGE", "EXPECT_INHERIT_10K_PLUS", "FIN_HEALTH.0", "FIN_HEALTH.1")
colnames(treats) <- c("RACE_ETHNICITY", "HIGHEST_EDUCATION_OF_RAISERS", "BINARIZED_GENDER","LAYOFF_PANDEMIC", "STATE", "EDUCATION_LEVEL", "NUM_DEPENDENT_CHILDREN", "AGE", "EXPECT_INHERIT_10K_PLUS", "FIN_HEALTH.0", "FIN_HEALTH.1")


mu_hat_1 = lm(FIN_HEALTH.1 ~ RACE_ETHNICITY + HIGHEST_EDUCATION_OF_RAISERS + BINARIZED_GENDER + LAYOFF_PANDEMIC + STATE + EDUCATION_LEVEL + NUM_DEPENDENT_CHILDREN + AGE + EXPECT_INHERIT_10K_PLUS, data=treats)

mu_hat_0 = lm(FIN_HEALTH.0 ~ RACE_ETHNICITY + HIGHEST_EDUCATION_OF_RAISERS + BINARIZED_GENDER + LAYOFF_PANDEMIC + STATE + EDUCATION_LEVEL + NUM_DEPENDENT_CHILDREN + AGE + EXPECT_INHERIT_10K_PLUS, data=controls)

mu_hat01 <- predict(mu_hat_0, treats)

mu_hat00 <- predict(mu_hat_0, controls)
mu_hat11 <- predict(mu_hat_1, treats)

bias <- mean(mu_hat01 - mu_hat00)

bias_adjusted_tau <- test_stat - bias


v_1 = 1/(n^2) * sum( (summary_2$FIN_HEALTH.1 - mu_hat11 )^2)
v_0 = 1/(n^2) * sum( (summary_2$FIN_HEALTH.0 - mu_hat00 )^2)

var_hat <- v_1 + v_0

sprintf("Bias_Adjusted Tau: %f", bias_adjusted_tau)
sprintf("Variance Estimate: %f", var_hat)

ci_lower <- bias_adjusted_tau - 1.96*((var_hat)^(1/2))
ci_upper <- bias_adjusted_tau + 1.96*((var_hat)^(1/2))
sprintf("Confidence interval is: [%f, %f]",ci_lower, ci_upper )



```








